\section{Temporal-Causal Modality: Consistency Validation Through Light Propagation Constraints}
\label{sec:temporal}

\subsection{Causal Structure of Light Propagation}

Light propagation obeys causality: photon at position $\mathbf{r}$ and time $t$ can only have originated from sources within past light cone. Causal Green's function for electromagnetic field \cite{Jackson1999}:
\begin{equation}
G(\mathbf{r}, t; \mathbf{r}', t') = \frac{\delta(t - t' - |\mathbf{r} - \mathbf{r}'|/c)}{4\pi|\mathbf{r} - \mathbf{r}'|} \Theta(t - t')
\end{equation}
where $\Theta$ is Heaviside step function enforcing $t > t'$ (causality) and $\delta$ function enforces light-speed propagation.

Field at $(\mathbf{r}, t)$ from source distribution $\rho(\mathbf{r}', t')$:
\begin{equation}
\Phi(\mathbf{r}, t) = \int_{-\infty}^t dt' \int_V d^3\mathbf{r}' \, \rho(\mathbf{r}', t') G(\mathbf{r}, t; \mathbf{r}', t')
\end{equation}

Causality constraint: field depends only on sources at retarded times $t' = t - |\mathbf{r} - \mathbf{r}'|/c$.

\subsection{Structural Prediction and Light Distribution Consistency}

Given structure $S(t_0)$ measured at time $t_0$, predict structure $\hat{S}(t_1)$ at future time $t_1 > t_0$ using physical constraints (conservation laws, thermodynamic evolution, reaction kinetics).

Predicted structure implies light distribution:
\begin{equation}
L^{\text{pred}}(\mathbf{r}, t_1) = \int_{t_0}^{t_1} dt' \int_V d^3\mathbf{r}' \, \sigma(\mathbf{r}', t'; \hat{S}) G(\mathbf{r}, t_1; \mathbf{r}', t')
\end{equation}
where $\sigma(\mathbf{r}', t'; \hat{S})$ is light source/scattering distribution implied by predicted structure $\hat{S}$.

\begin{theorem}[Temporal-Causal Consistency]
\label{thm:temporal_causal}
Predicted structure $\hat{S}(t_1)$ is consistent with physics if and only if predicted light distribution matches observed distribution:
\begin{equation}
L^{\text{pred}}(\mathbf{r}, t_1) = L^{\text{obs}}(\mathbf{r}, t_1)
\end{equation}
within measurement uncertainty $\delta L$.
\end{theorem}

\begin{proof}
Light propagation is deterministic: given source distribution $\rho(\mathbf{r}', t')$, field $\Phi(\mathbf{r}, t)$ is uniquely determined by Maxwell's equations with causal boundary conditions.

Structure determines source distribution through reflection, scattering, and emission. If predicted structure is correct, implied source distribution equals actual source distribution, thus predicted light distribution equals observed distribution.

Conversely, if light distributions differ by more than measurement uncertainty, implied source distributions differ, thus predicted structure is incorrect.
\end{proof}

\subsection{Exclusion Through Causal Inconsistency}

\begin{corollary}[Causal Exclusion]
\label{cor:causal_exclusion}
Candidate structures predicting light distributions violating causal consistency:
\begin{equation}
\int |\mathbf{L}^{\text{pred}}(\mathbf{r}, t) - \mathbf{L}^{\text{obs}}(\mathbf{r}, t)|^2 d^3\mathbf{r} > (\delta L)^2 V
\end{equation}
are excluded, where $V$ is observation volume.
\end{corollary}

\begin{proof}
Follows directly from Theorem~\ref{thm:temporal_causal}. Structures producing light distributions outside measurement uncertainty violate causal consistency and are excluded.
\end{proof}

Exclusion factor depends on precision of light measurement and complexity of prediction. For optical measurements with signal-to-noise ratio SNR $\sim 10^3$ over volume with $\sim 10^6$ resolvable elements:
\begin{equation}
\epsilon_{\text{causal}} \sim \text{SNR}^{-1} \sim 10^{-3}
\end{equation}
per spatial element.

Multiple measurements at different times $\{t_j\}$ compound exclusion:
\begin{equation}
\epsilon_{\text{causal}}^{N_t} \sim (10^{-3})^{N_t}
\end{equation}

For $N_t = 5$ temporal samples, total exclusion $\sim 10^{-15}$.

\subsection{Computational Implementation}

Computing predicted light distribution requires:

\textbf{Step 1}: Given structure $S(t_0)$, extract source distribution $\rho(\mathbf{r}, t_0)$ (includes scatterers, fluorophores, reflection boundaries).

\textbf{Step 2}: Evolve structure forward in time using physical laws:
\begin{equation}
\frac{d S}{dt} = F(S, T, P, \ldots)
\end{equation}
where $F$ encodes dynamics (molecular motion, reactions, diffusion).

\textbf{Step 3}: Compute implied source distribution at time $t_1$:
\begin{equation}
\rho(\mathbf{r}, t_1) = \sigma(\mathbf{r}; S(t_1))
\end{equation}

\textbf{Step 4}: Integrate causal Green's function:
\begin{equation}
L^{\text{pred}}(\mathbf{r}, t_1) = \int_{t_0}^{t_1} dt' \int_V d^3\mathbf{r}' \, \rho(\mathbf{r}', t') G(\mathbf{r}, t_1; \mathbf{r}', t')
\end{equation}

\textbf{Step 5}: Compare to observed $L^{\text{obs}}(\mathbf{r}, t_1)$. If mismatch exceeds $\delta L$, structure prediction is incorrect.

This provides self-consistency check without requiring explicit light propagation simulation—only verification that predicted structure produces observed light distribution.

\subsection{Multi-Time Validation}

Measuring at multiple times $\{t_j\}_{j=0}^{N_t}$ enables trajectory validation. Predicted trajectory $\{S(t_j)\}$ must satisfy:
\begin{equation}
L^{\text{pred}}(\mathbf{r}, t_j) = L^{\text{obs}}(\mathbf{r}, t_j) \quad \forall j
\end{equation}

This overdetermines structure: $N_t$ measurements provide $N_t$ constraints, while structure evolution depends on continuous parameters. Inconsistency indicates error in structural prediction.

\begin{proposition}[Temporal Overdetermination]
\label{prop:temporal_overdetermination}
For structure with $N_{\text{param}}$ continuous parameters, $N_t > N_{\text{param}}$ temporal measurements overdetermine structure, enabling error detection through consistency checking.
\end{proposition}

\begin{proof}
$N_{\text{param}}$ parameters require $N_{\text{param}}$ independent measurements for unique determination. Additional measurements ($N_t > N_{\text{param}}$) provide redundancy. If all measurements are consistent with single parameter set, structure is validated. If inconsistent, at least one measurement or structural prediction is erroneous.
\end{proof}

\subsection{Relationship to Retarded Potentials}

Standard electromagnetic theory uses retarded potentials \cite{Jackson1999}:
\begin{equation}
\mathbf{A}(\mathbf{r}, t) = \frac{\mu_0}{4\pi} \int \frac{\mathbf{J}(\mathbf{r}', t - |\mathbf{r} - \mathbf{r}'|/c)}{|\mathbf{r} - \mathbf{r}'|} d^3\mathbf{r}'
\end{equation}

Current at retarded time $t_{\text{ret}} = t - |\mathbf{r} - \mathbf{r}'|/c$ determines field at $(\mathbf{r}, t)$. This is identical to causal Green's function formulation.

Our contribution is recognizing this causal structure provides constraint for structural validation: predicted structures must produce retarded potentials consistent with observation.

