\section{Entropy Equivalence}
\label{sec:entropy-equivalence}

\subsection{Three Routes to Entropy}

We establish that three apparently distinct processes---oscillation, categorization, and partitioning---yield identical entropy formulas.

\begin{theorem}[Oscillation-Category-Partition Equivalence]
\label{thm:equivalence}
For a system with $M$ independent subsystems, each admitting $n$ distinguishable states, the following three entropy calculations yield identical results:
\begin{enumerate}
    \item Oscillation entropy: $S_{\text{osc}} = k_B M \ln n$
    \item Categorical entropy: $S_{\text{cat}} = k_B M \ln n$
    \item Partition entropy: $S_{\text{part}} = k_B M \ln n$
\end{enumerate}
\end{theorem}

\subsection{Oscillation Entropy}

\begin{definition}[Oscillator Phase Space]
An oscillator with frequency $\omega$ confined to energy $E$ occupies phase space volume
\begin{equation}
    \Gamma = \frac{2\pi E}{\omega} = \frac{E}{\nu}
\end{equation}
where $\nu = \omega/2\pi$ is the frequency.
\end{definition}

\begin{proposition}
For $M$ independent oscillators, each with $n$ accessible energy levels, the phase space volume is $\Gamma = n^M$ in units of $h^M$. The Boltzmann entropy is
\begin{equation}
    S_{\text{osc}} = k_B \ln \Gamma = k_B M \ln n
\end{equation}
\end{proposition}

\begin{proof}
Each oscillator independently accesses $n$ levels, giving $n^M$ total microstates. Direct application of $S = k_B \ln W$ yields the result.
\end{proof}

\subsection{Categorical Entropy}

\begin{definition}[Categorical State Space]
A category $\mathcal{C}$ with $n$ objects admits $n$ morphisms from any object to itself (assuming skeletal category). The categorical state space is the set of all possible object assignments.
\end{definition}

\begin{proposition}
For $M$ independent categorical variables, each selecting from $n$ categories, the total number of configurations is $n^M$. The categorical entropy is
\begin{equation}
    S_{\text{cat}} = k_B \ln(n^M) = k_B M \ln n
\end{equation}
\end{proposition}

In mass spectrometry, categories correspond to chemical identities. A fragmentation pattern assigns each fragment to a category (chemical structure). The categorical entropy measures the information content of the assignment.

\subsection{Partition Entropy}

\begin{definition}[Partition Function]
A partition of set $X$ with $M$ elements into $n$ non-empty subsets has multiplicity
\begin{equation}
    W = \frac{M!}{\prod_{i=1}^n m_i!}
\end{equation}
where $m_i$ is the size of subset $i$.
\end{definition}

For equal partitions ($m_i = M/n$ for all $i$), the Stirling approximation yields:

\begin{proposition}
The partition entropy for $M$ elements distributed equally among $n$ subsets is
\begin{equation}
    S_{\text{part}} = k_B \ln W \approx k_B M \ln n - k_B n \ln(M/n) \approx k_B M \ln n
\end{equation}
in the limit $M \gg n$.
\end{proposition}

\subsection{Equivalence Proof}

\begin{proof}[Proof of Theorem~\ref{thm:equivalence}]
All three entropies reduce to $k_B M \ln n$:
\begin{itemize}
    \item Oscillation: $M$ oscillators, each accessing $n$ states, give $n^M$ configurations.
    \item Category: $M$ categorical variables, each with $n$ values, give $n^M$ assignments.
    \item Partition: $M$ elements partitioned into $n$ equal subsets give (in leading order) $n^M$ arrangements.
\end{itemize}
The mathematical structure is identical: the logarithm of a counting function over $M$ independent choices from $n$ options.
\end{proof}

\subsection{Physical Interpretation}

The equivalence theorem implies that entropy is fundamentally about distinguishability, not about the physical substrate generating the distinctions. Whether distinctions arise from oscillator energy levels, categorical assignments, or set partitions, the entropy formula is invariant.

For mass spectrometry, this equivalence has practical consequences:

\begin{corollary}[Hardware Independence of Entropy]
The entropy of a molecular fragmentation pattern is independent of the mass analyzer used to measure it, provided all analyzers resolve the same number of distinct states.
\end{corollary}

A quadrupole and an Orbitrap measuring the same fragmentation at the same mass resolution must yield identical entropy values, despite operating at different frequencies, different pressures, and different detection mechanisms.

